---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To read the associated Excel-files the following function is used. Values below the detection limit are replaced by half of the detection limit (not applicable for TOC), decimal comma is replaced by decimal point.

```{r}
import_slu_mvm_data_excel <- function(filename, numeric_var = NA, bad_quality_na = TRUE, sheet = 2){ # function for importing excel files into R
  #numeric var: the first column with measured numeric variables
  replace_less_than_var <- function(values, bad_quality_na){ # function for replacing values at the detection limit with half values
    values_parsed <- values %>% as.character() %>% parse_number(locale = locale(decimal_mark = ","))

   which_intervals <- which(substring(values,1,1) == "[")
   which_less_than <- which(substring(values,1,1) == "<")
   if (length(which_less_than) > 0) {
      values_less_than_half <- values[which_less_than] %>% gsub(",", ".", .) %>% gsub("<","",.) %>% as.numeric()
      values_parsed[which_less_than] <- values_less_than_half/2}

   if (bad_quality_na == TRUE) {values_parsed[which_intervals] <- NA}
   else{
      values_intervals <- values[which_intervals] %>%
        gsub("\\[","", .) %>%
        gsub("\\]","", .) %>%
        gsub(",",".", .) %>%
        as.numeric()
      values_parsed[which_intervals] <- values_intervals}

   return(values_parsed)
  }

  if (is.na(numeric_var) == T) {stop("Supply the column index of first measured numeric variable")}
out <- suppressWarnings(read_excel(filename, sheet = sheet,guess_max = 50000))
out <-
  
  mutate_at(out, c(numeric_var:ncol(out)),
            replace_less_than_var,
            bad_quality_na = bad_quality_na)
out <-
  mutate_at(out, c(1:(numeric_var - 1)),
            (function(x) {
                          x %>%
                            as.character() %>%
                            parse_guess(locale = locale(decimal_mark = ","))
              })
           )

}
```


Load needed libraries
```{r}
library(mgcv)
library(tidyverse)
library(readxl)
library(lubridate)
library(janitor)
library(GWmodel)
library(knitr)
library(ape)
```

Prepare data
############


Read data from the SLS, remove lakes situated on Gotland, define year as the year of observation unless the month of observation is January, then the observation is counted for the year before. TOC is log-transformed. Include a variable to identify these lakes as SLS lakes
```{r}
lakes <- import_slu_mvm_data_excel("SLS.xlsx", numeric_var = 29)

lakes%>%mutate(year1=year(Provdatum), 
               month=month(Provdatum),
               log_toc=log10(`TOC (mg/l C)`), 
               toc=`TOC (mg/l C)`,
               year=case_when(month==1~year1-1,
                              TRUE~year1),
               SLS=1)%>%
  filter(year!=2007,(`Stationskoordinat N/X`>6445896 | `Stationskoordinat E/Y`<686982.1))%>%
  clean_names()%>%drop_na(toc)->lakes1
```

Read data from the STL, remove lakes situated on Gotland,extract observations from September to November. TOC is log-transformed.Compute means in case there are several observations per station and year. This file contains data from 2012 to 2021.
```{r}
trendlakes<- import_slu_mvm_data_excel("STL2012-2021.xlsx", numeric_var = 26)%>%
  clean_names()%>%
  filter(provtagningsmanad>8 & provtagningsmanad <12 & provtagningsar>=2007, (stationskoordinat_n_x>6445896 | stationskoordinat_e_y<686982.1), max_provdjup_m<=1)%>%
  mutate(year=provtagningsar, log_toc=log10(toc_mg_l_c), sls=0)%>%
  group_by(nationellt_overvakningsstations_id, stationskoordinat_e_y, stationskoordinat_n_x, overvakningsstation,  year, sls)%>%
  summarize(mlog_toc=mean(log_toc, na.rm=TRUE), mtoc=mean(toc_mg_l_c, na.rm=TRUE))%>%rename(log_toc=mlog_toc, toc=mtoc, ID=nationellt_overvakningsstations_id)
```

Read a similar file with data before 2012.
```{r}
trendlakes2<- import_slu_mvm_data_excel("STL2007-2011.xlsx", numeric_var = 26)%>%
  clean_names()%>%
  filter(provtagningsmanad>8 & provtagningsmanad <12 & provtagningsar>2007, ( stationskoordinat_n_x>6445896 | stationskoordinat_e_y<686982.1), max_provdjup_m<=1)%>%
  mutate(year=provtagningsar, log_toc=log10(toc_mg_l_c), sls=0)%>%
  group_by(nationellt_overvakningsstations_id, stationskoordinat_e_y, stationskoordinat_n_x, overvakningsstation,  year, sls)%>%
  summarize(mlog_toc=mean(log_toc, na.rm=TRUE), mtoc=mean(toc_mg_l_c, na.rm=TRUE))%>%rename(log_toc=mlog_toc, toc=mtoc, ID=nationellt_overvakningsstations_id)


```

Read coordinates that define the border of Sweden, including the largest lakes.
```{r}
border <- read_excel("sverige_smooth_coordinates_SWEREF.xlsx")
border1<-subset(border, namn=="sverige", select=c(East_SWERE, North_SWER))
border_vänern<-subset(border, namn=="vänern", select=c(East_SWERE, North_SWER))
border_vättern<-subset(border, namn=="vättern", select=c(East_SWERE, North_SWER))
border_mälaren<-subset(border, namn=="mälaren", select=c(East_SWERE, North_SWER))

```


Set unique IDs for all lakes in the SLS. If present the national station-ID is used, otherwise the MVM-ID.
```{r}

#Obtain nationellt_overvakningsstations_id from koordinates
lakes1%>%dplyr::select(stationskoordinat_e_y, stationskoordinat_n_x, nationellt_overvakningsstations_id)%>%group_by(stationskoordinat_n_x, stationskoordinat_e_y)%>%distinct()%>%slice(1)%>%filter(!is.na(nationellt_overvakningsstations_id))%>%
  rename(new_ovv=nationellt_overvakningsstations_id)->stations1

#Join with original dataset to get the same ID for same koordinates
lakes1%>%select(-nationellt_overvakningsstations_id)%>%left_join(stations1)%>%
  mutate(ID=new_ovv)->lakes_a1

#Obtain mvm_id from koordinates
lakes_a1%>%dplyr::select(stationskoordinat_e_y, stationskoordinat_n_x, md_mvm_id)%>%group_by(stationskoordinat_n_x, stationskoordinat_e_y)%>%distinct()%>%slice(1)%>%filter(!is.na(md_mvm_id))%>%
  rename(new_ovv=md_mvm_id)->stations2

#Join with dataset to fill ID for stations that do not have a national ID
lakes_a1%>% select(-new_ovv)%>%left_join(stations2)%>%
mutate(ID=case_when(is.na(ID)~as.character(new_ovv),
                      TRUE~ID))->lakes_a


```
Identify stations with only a single observations and remove them
```{r}
lakes_a%>%group_by(ID, stationskoordinat_e_y, stationskoordinat_n_x)%>% mutate(n=n())%>%filter(n<=1)%>%arrange(ID)%>%
  dplyr::select( stationskoordinat_e_y, stationskoordinat_n_x, overvakningsstation, ID)->only_single


lakes_a%>%filter(!ID %in% only_single$ID) ->lakes2
```

Merge SLS and STL data
```{r}
lakes2%>%select(ID, overvakningsstation, stationskoordinat_e_y, stationskoordinat_n_x,  year, sls, log_toc, toc)%>%
  full_join(trendlakes)%>%
  full_join(trendlakes2) ->lakes_joint

```

The station-wise mean-centered TOC variable is computed. 

Create a dataset with all data and one that only contains SLS data, 
```{r}
lakes_joint%>% group_by(ID, stationskoordinat_e_y, stationskoordinat_n_x)%>%
  summarize(sd_toc=sd(log_toc), mean_toc=mean(log_toc, na.rm=TRUE))->lakes_sd

lakes_joint%>%
  left_join(lakes_sd)%>%
  mutate(cent_toc=log_toc-mean_toc)%>%
  drop_na(log_toc)->lakes_all

lakes_all%>%filter(sls==1)->lakes_SLS

```


Bandwidth search. Takes time, gives 400.
```{r}
#lakes_all.spdf <- SpatialPointsDataFrame(lakes_all[, c(3,4)], lakes_all)
#bw<-bw.gwr(cent_toc ~ year,adaptive=TRUE,kernel="bisquare", parallel.method=TRUE,data=lakes_all.spdf)
```

Run model for the combined dataset.
###################################


```{r}
lakes_all.spdf <- SpatialPointsDataFrame(lakes_all[, c(3,4)], lakes_all)
model_toc_SLS_and_STL<-gwr.basic(cent_toc ~ year,bw=400,adaptive=TRUE,kernel="bisquare", parallel.method=TRUE,data=lakes_all.spdf)

pval_toc_SLS_and_STL<-gwr.t.adjust(model_toc_SLS_and_STL)


```

Figure 5, left: Slope coefficients from the model based on both the SLS and STL
```{r}
TOC_SLR<-ggplot(lakes_all, aes(x=stationskoordinat_e_y,y=stationskoordinat_n_x))+
  geom_polygon(data = border1,aes(x = East_SWERE, y = North_SWER),fill = "gray",alpha = 0.25)+
  geom_polygon(data = border_vänern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_vättern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_mälaren,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_point(aes(colour=model_toc_SLS_and_STL$SDF$year))+
  geom_path(data = border1,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vänern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vättern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_mälaren,aes(x = East_SWERE, y = North_SWER))+
  coord_fixed()+
  scale_colour_gradient2(low ="darkblue",mid="white", high = "darkred", midpoint=0, limits=c(-0.038, 0.038), space = "rgb", na.value = "grey50", guide = "colourbar", guide_legend(title="Linear trend slope"))+
  theme_classic()+
  ylab("")+
  xlab("")
```

Significant trends for model based on both the SLS and STL
```{r}
lakes_all%>%mutate(pval0=case_when(pval_toc_SLS_and_STL$SDF$year_p<0.05 & model_toc_SLS_and_STL$SDF$year>0~1,
                              pval_toc_SLS_and_STL$SDF$year_p<0.05 & model_toc_SLS_and_STL$SDF$year<0 ~-1,
                              TRUE~0))%>%
  group_by(stationskoordinat_e_y, stationskoordinat_n_x)%>%slice(1)->lakes_p_all



ggplot(lakes_p_all, aes(x=stationskoordinat_e_y,y=stationskoordinat_n_x))+
  geom_polygon(data = border1,aes(x = East_SWERE, y = North_SWER),fill = "gray",alpha = 0.25)+
  geom_polygon(data = border_vänern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_vättern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_mälaren,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_point(aes(colour=as.factor(pval0)))+
  geom_path(data = border1,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vänern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vättern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_mälaren,aes(x = East_SWERE, y = North_SWER))+
  coord_fixed()+
  scale_colour_manual(values=c("blue", "gray", "red"))+
  theme_classic()+
  theme(legend.position="None")+
  ylab("")+
  xlab("")
```

```{r}
SEplot_SLS_STL<-ggplot(lakes_all, aes(x=stationskoordinat_e_y,y=stationskoordinat_n_x))+
  geom_polygon(data = border1,aes(x = East_SWERE, y = North_SWER),fill = "gray",alpha = 0.25)+
  geom_polygon(data = border_vänern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_vättern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_mälaren,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_point(aes(colour=model_toc_SLS_and_STL$SDF$year_SE))+
  geom_path(data = border1,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vänern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vättern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_mälaren,aes(x = East_SWERE, y = North_SWER))+
  coord_fixed()+
  scale_colour_gradient2(low ="orange",mid="darkolivegreen1", high = "darkgreen", midpoint=0.001, space = "rgb", limits=c(0.0008, 0.0013), na.value = "grey50", guide = "colourbar", guide_legend(title="Standard error of slope"))+
  theme_classic()+
  ylab("")+
  xlab("")
```




Differences in results between the SLS+STL model and the SLS alone model
########################################################################

Run model for only SLS 
```{r}

lakes_SLS.spdf <- SpatialPointsDataFrame(lakes_SLS[, c(3,4)], lakes_SLS)

model_toc_SLS<-gwr.basic(cent_toc ~ year,bw=370,adaptive=TRUE,kernel="bisquare", parallel.method=TRUE,data=lakes_SLS.spdf)

pval_toc_SLS<-gwr.t.adjust(model_toc_SLS)


```



Differences in slopes between models
```{r}
lakes_all%>%mutate(coef_all=model_toc_SLS_and_STL$SDF$year,
                pval_all=pval_toc_SLS_and_STL$SDF$year_p)->lakes_all1

lakes_SLS%>%mutate(coef_SLS=model_toc_SLS$SDF$year,
                  pval_SLS=pval_toc_SLS$SDF$year_p)%>%right_join(lakes_all1, by=c("ID","year"))%>%
  mutate(est_diff=coef_all-coef_SLS)%>%
  arrange(desc(abs(est_diff)))%>%
  select(ID, overvakningsstation.x,sls.x, sls.y, est_diff, log_toc.x, year, coef_all, coef_SLS, stationskoordinat_e_y.x, stationskoordinat_n_x.x)->diffs
```

Figure 5, right: Differences in estimated slope coefficients for the model using SLS and STL compared to SLS only.
```{r}
library(ggforce)
TOC_changes<-ggplot(diffs, aes(x=stationskoordinat_e_y.x,y=stationskoordinat_n_x.x))+
  geom_polygon(data = border1,aes(x = East_SWERE, y = North_SWER),fill = "gray",alpha = 0.25)+
  geom_polygon(data = border_vänern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_vättern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_mälaren,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_point(aes(colour=est_diff))+
  geom_path(data = border1,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vänern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vättern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_mälaren,aes(x = East_SWERE, y = North_SWER))+
  coord_fixed()+
  scale_colour_gradient2(low ="blue",mid="white", high = "red", midpoint=0, space = "rgb", na.value = "grey50", guide = "colourbar", guide_legend(title="Difference in slope"))+
  geom_point(data=lakes_all%>%filter(sls==0), aes(x=stationskoordinat_e_y, y=stationskoordinat_n_x), col="black", lwd=1)+
  theme_classic()+
  ylab("")+
  xlab("")
```

```{r}
ggarrange(TOC_SLR, TOC_changes, ncol=2, nrow=1, widths=c(550,550), heights=c(700, 700))

```


Figure S3 right: Differences in significance for models based on SLS and STL compared to SLS only
```{r}
diffs%>%group_by(ID)%>%slice(1)%>%ungroup()%>%mutate(ind=case_when(est_diff>0.001 & coef_SLS>0~"increased positive",
                          est_diff>0.001 & coef_SLS<0~"decreased negative",
                          est_diff< -0.001 & coef_SLS>0~"decreased positive",
                          est_diff< -0.001 & coef_SLS<0 ~"increased negative",
                          est_diff== 0~"no change",
                          coef_SLS==0~"coefSLS 0"))->diffs1

SLS_STL_changes<-ggplot(diffs1, aes(x=stationskoordinat_e_y.x,y=stationskoordinat_n_x.x))+
  geom_polygon(data = border1,aes(x = East_SWERE, y = North_SWER),fill = "gray",alpha = 0.25)+
  geom_polygon(data = border_vänern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_vättern,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_polygon(data = border_mälaren,aes(x = East_SWERE, y = North_SWER),fill = "lightblue",alpha = 0.5)+
  geom_point(aes(colour=ind))+
  scale_color_manual(values = c("increased positive" = "red",
                                "increased negative"="blue",
                                "decreased positive"="orange",
                                "decreased negative"="turquoise"),
                     labels = c("increased positive", "increased negative", "decreased positive", "decreased negative"),
                     name="Change", na.value="gray80")+
  geom_path(data = border1,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vänern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_vättern,aes(x = East_SWERE, y = North_SWER))+
  geom_path(data = border_mälaren,aes(x = East_SWERE, y = North_SWER))+
  coord_fixed()+
  theme_classic()+
  ylab("")+
  xlab("")
```


Comparison of trends in SLS and STL for specific identified areas
#################################################################

Figure 6, upper left: Lake Kupesjön
```{r}
distance<-lakes_all%>%mutate(dist=(stationskoordinat_e_y-450408)^2 +(stationskoordinat_n_x-6331514)^2)%>%group_by(ID)%>%slice(1)%>%ungroup()%>%arrange(dist)%>%slice(30)

Fig6_ul<-lakes_all%>%mutate(dist=(stationskoordinat_e_y-450408)^2 +(stationskoordinat_n_x-6331514)^2)%>%filter(dist <= distance$dist & sls==1)%>%
  mutate(ID_new=paste(ID, sls))%>%
  ggplot()+
  geom_point(aes(x=year, y=log_toc), col="gray40")+
  geom_line(aes(x=year, y=log_toc, group=ID), col="gray40")+
  theme(legend.position = "none")+
  geom_line(data=lakes_all%>%mutate(dist=(stationskoordinat_e_y-450408)^2 +(stationskoordinat_n_x-6331514)^2)%>%filter(dist <= distance$dist & sls==0),
             aes(x=year, y=log_toc, group=overvakningsstation, lty=overvakningsstation), lwd=1.1) +
  theme_classic()+
  ylab("TOC (log-transformed)")+
  xlab("Year")+
  theme(legend.position="bottom")+
  theme(text= element_text(size=16))+
  guides(linetype=guide_legend(title="Station"))

```

Figure 6, upper right: Älvasjön
```{r}
distance<-lakes_all%>%mutate(dist=(stationskoordinat_e_y-384420)^2 +(stationskoordinat_n_x-6284776)^2)%>%group_by(ID)%>%slice(1)%>%ungroup()%>%arrange(dist)%>%slice(30)


Fig6_ur<-lakes_all%>%mutate(dist=(stationskoordinat_e_y-384420)^2 +(stationskoordinat_n_x-6284776)^2)%>%filter(dist <= distance$dist & sls==1)%>%
  mutate(ID_new=paste(ID, sls))%>%
  ggplot()+
  geom_point(aes(x=year, y=log_toc), col="gray40")+
  geom_line(aes(x=year, y=log_toc, group=ID), col="gray40")+
  theme(legend.position = "none")+
  geom_line(data=lakes_all%>%mutate(dist=(stationskoordinat_e_y-384420)^2 +(stationskoordinat_n_x-6284776)^2)%>%filter(dist <= distance$dist & sls==0),
             aes(x=year, y=log_toc, group=overvakningsstation, lty=overvakningsstation), lwd=1.1) +
  theme_classic()+
  ylab("TOC (log-transformed)")+
  xlab("Year")+
  theme(legend.position="bottom")+
  theme(text= element_text(size=16))+
  guides(linetype=guide_legend(title="Station"))
```


Figure 6, lower left: Norasjön
```{r}

distance<-lakes_all%>%mutate(dist=(stationskoordinat_e_y-647006)^2 +(stationskoordinat_n_x-6538408)^2)%>%group_by(ID)%>%slice(1)%>%ungroup()%>%arrange(dist)%>%slice(30)

Fig6_ll<-lakes_all%>%mutate(dist=(stationskoordinat_e_y-647006)^2 +(stationskoordinat_n_x-6538408)^2)%>%filter(dist <= distance$dist & sls==1)%>%
  mutate(ID_new=paste(ID, sls))%>%
  ggplot()+
  geom_point(aes(x=year, y=log_toc), col="gray40")+
  geom_line(aes(x=year, y=log_toc, group=ID), col="gray40")+
  theme(legend.position = "none")+
  geom_line(data=lakes_all%>%mutate(dist=(stationskoordinat_e_y-647006)^2 +(stationskoordinat_n_x-6538408)^2)%>%filter(dist <= distance$dist & sls==0),
             aes(x=year, y=log_toc, group=overvakningsstation, lty=overvakningsstation), lwd=1.1) +
  theme_classic()+
  ylab("TOC (log-transformed)")+
  xlab("Year")+
  theme(legend.position="bottom")+
  theme(text= element_text(size=16))+
  guides(linetype=guide_legend(title="Station"))

```


Figure 6, lower right: Outside Karlshamn
```{r}
distance<-lakes_all%>%mutate(dist=(stationskoordinat_e_y-496102)^2 +(stationskoordinat_n_x-6225833)^2)%>%group_by(ID)%>%slice(1)%>%ungroup()%>%arrange(dist)%>%slice(30)


Fig6_lr<-lakes_all%>%mutate(dist=(stationskoordinat_e_y-496102)^2 +(stationskoordinat_n_x-6225833)^2)%>%filter(dist <= distance$dist & sls==1)%>%
  mutate(ID_new=paste(ID, sls))%>%
  ggplot()+
  geom_point(aes(x=year, y=log_toc), col="gray40")+
  geom_line(aes(x=year, y=log_toc, group=ID), col="gray40")+
  theme(legend.position = "none")+
  geom_line(data=lakes_all%>%mutate(dist=(stationskoordinat_e_y-496102)^2 +(stationskoordinat_n_x-6225833)^2)%>%filter(dist <= distance$dist & sls==0),
             aes(x=year, y=log_toc, group=overvakningsstation, lty=overvakningsstation), lwd=1.1) +
  theme_classic()+
  ylab("TOC (log-transformed)")+
  xlab("Year")+
  theme(legend.position="bottom")+
  theme(text= element_text(size=16))+
  guides(linetype=guide_legend(title="Station"))
```




Area between Vänern and Vättern.
```{r}
distance<-lakes_all%>%mutate(dist=(stationskoordinat_e_y-416952)^2 +(stationskoordinat_n_x-6479250)^2)%>%group_by(ID)%>%slice(1)%>%ungroup()%>%arrange(dist)%>%slice(30)


lakes_all%>%mutate(dist=(stationskoordinat_e_y-416952)^2 +(stationskoordinat_n_x-6479250)^2)%>%filter(dist <= distance$dist & sls==1)%>%
  mutate(ID_new=paste(ID, sls))%>%
  ggplot()+
  geom_point(aes(x=year, y=log_toc), col="gray40")+
  geom_line(aes(x=year, y=log_toc, group=ID), col="gray40")+
  theme(legend.position = "none")+
  geom_line(data=lakes_all%>%mutate(dist=(stationskoordinat_e_y-416952)^2 +(stationskoordinat_n_x-6479250)^2)%>%filter(dist <= distance$dist & sls==0),
             aes(x=year, y=log_toc, group=overvakningsstation, lty=overvakningsstation), lwd=1.1) +
  theme_classic()+
  ylab("TOC (log-transformed)")+
  xlab("Year")+
  theme(legend.position="bottom")+
  theme(text= element_text(size=16))+
  guides(linetype=guide_legend(title="Station"))
```


Nonlinear trends for the STL Program using generalized additive models
######################################################################

```{r}
source("sourcecode_screening2020_dummy2.R")
```

Read data and combine into one dataset
```{r}
trendlakes2012<- import_slu_mvm_data_excel("STL2012-2021.xlsx", numeric_var = 26)%>%
  clean_names()%>%
  filter(provtagningsar>=2011, max_provdjup_m<=1)%>%
  mutate(year=provtagningsar, log_toc=log10(toc_mg_l_c), sls=0, month=provtagningsmanad)%>%
  rename(ID=nationellt_overvakningsstations_id)
  

trendlakes2008<- import_slu_mvm_data_excel("STL2007-2011.xlsx", numeric_var = 26)%>%
  clean_names()%>%
  filter(provtagningsar>=2007, max_provdjup_m<=1)%>%
  mutate(year=provtagningsar, log_toc=log10(toc_mg_l_c), sls=0, month=provtagningsmanad)%>%
  rename(ID=nationellt_overvakningsstations_id)
  

trendlakes_all<-trendlakes2012%>%full_join(trendlakes2008)

```

Create seasonal dummy
```{r}
trendlakes_all%>%
  mutate(overvakningsstation=case_when(overvakningsstation=="Stora Skärsjön" & ID=="00261328"~"Stora Skärsjön 2",
                                                 overvakningsstation=="Skärgölen" & ID=="00260889"~"Skärgölen 2",
                                                 TRUE~overvakningsstation),
                   seas=case_when(month>2 & month <6~2,
                                  month>5 & month <9~3,
                                  month>8 & month< 12~4,
                                  TRUE~1),
                   date=ymd(make_date(year, month, 1)))%>%
  drop_na(log_toc)%>%
  filter(!overvakningsstation %in% c("Älgsjön", "Fräcksjön", "Stora Gryten"), year<2022)->trendlakes_seas
```

Run screening models 
```{r}
trendlakes_seas%>%
   select(overvakningsstation, 
         date, 
         seas,
         log_toc,
         )%>%
  mutate(SiteID=as.factor(overvakningsstation))%>%
  pivot_longer(cols=c("log_toc")) %>% 
    screeningmodeling(values=value,
                    datevar = date, 
                    seas=seas,
                    #seasonal = F, 
                    link = "identity", 
                    conf.type = "conf",
                    beep = TRUE, 
                    tdist = F,
                    autocor = TRUE,
                    overvakningsstation,
                    name) ->
results 
```

Figure S5: Screening plot for nonlinear trends in TOC.
```{r}
results%>%left_join(trendlakes_seas%>%group_by(overvakningsstation)%>%slice(1)%>%select(overvakningsstation, stationskoordinat_e_y, stationskoordinat_n_x))->results1

results1 %>%plot_screeningtrends(y_id=overvakningsstation, sorting=desc(stationskoordinat_n_x))
plot_screening<-results1 %>%plot_screeningtrends_pvalues(y_id=overvakningsstation, sorting=desc(stationskoordinat_n_x))
results%>%filter(overvakningsstation=="Fjärasjö")%>%plot_individual_trend()

```


